---
title: 'Advanced Statistical Analysis and Modeling: Exploring Real-World Data with
  R'
author: "Ayobami Adeyeye"
date: "April 2021"
output:
  html_document: 
    keep_md: yes
  theme: united
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Executive Summary

The analysis of the USSeatBelts dataset aimed to understand the impact of various factors on road fatalities. The focus was on exploring the relationships between variables like year, miles driven, seat belt usage, income, and age with road fatalities.

# Introduction

**Background:** Road safety is a paramount concern globally, with significant implications for public health and policy. This study underscores the importance of statistical analysis in understanding and mitigating road fatalities.

**Problem Statement:** The primary inquiry of this case study is to determine the key factors influencing road fatalities, including the role of seat belt usage, and to predict their effects effectively.

## Data Overview:

**Data Source:** The 'USSeatBelts.csv' dataset, comprising various variables related to road safety.

```{r}
# data source https://vincentarelbundock.github.io/Rdatasets/datasets.html
# This code reads the 'USSeatBelts.csv' file into R as a dataframe named 'dataset'
dataset = read.csv("USSeatBelts.csv")

# Removes all rows with NA (missing) values.
dataset = na.omit(dataset)

#unction displays the first 5 rows of the dataset for a preliminary view.
head(dataset)

```

**Data Preparation:** The data preparation involved cleaning, handling missing values, and structuring the data for analysis.

```{r}
# Function provides a detailed summary of the dataset, including data types, missing values, and basic statistics
skimr::skim(dataset)

```

## Methodology:

**Statistical Techniques Used:** The study employed methods like GLM and correlation analysis to evaluate the data. The choice of these methods was based on their suitability for analyzing relationships within the dataset.

**Implementation in R:** The analysis was carried out in R, with a focus on leveraging its statistical capabilities for accurate data interpretation. Code snippets illustrate the application of these methods.


## Correlation Analysis:
```{r}
# creates a new dataframe 'num' with numeric columns from 'dataset
num=data.frame(dataset$year, dataset$miles, dataset$fatalities, dataset$seatbelt, dataset$income, dataset$age)

# calculates the correlation matrix of 'num' and stores it in 'ncor'
ncor=cor(num)

#  visualizes this correlation matrix.
library(corrplot)
corrplot(ncor)
ncor
```


+ **Seatbelt Usage and Fatalities:** A significant negative correlation of -0.4027138 was found between seat belt usage and road fatalities, indicating that increased seat belt usage is associated with fewer road deaths.  
+ **Income and Fatalities:** Income showed a strong negative correlation of -0.6928618 with fatalities, suggesting higher income levels may correlate with lower road fatalities.  
+ **Age and Fatalities:** The correlation between age and fatalities was -0.2730244, indicating a moderate negative relationship. 
+  **Year and Seatbelt Usage:** There was a strong positive correlation of 0.6792371 between the year and seat belt usage, implying an increase in seat belt usage over the years.  

```{r}
# prepares data for the GLM analysis by selecting relevant variables and creating a new dataframe 'df'
x1 = dataset$seatbelt
x2 = as.factor(dataset$speed70)
x3 = dataset$alcohol
y =  dataset$fatalities
df = data.frame(x1, x2, x3, y)

```

```{r,include=FALSE}
# Fits a GLM to the data in 'df' with 'y' as the response variable and other columns as predictors.
dataglm = glm(y~.,df,family = "gaussian")
#displays the results of the GLM.
summary(dataglm)
```

**Model trained using 80% of this dataset and suggest an appropriate GLM to model output to input variables.**

```{r}
# Sets a seed for reproducibility.
# Splits the data into training (80%) and testing (20%) sets.
set.seed(123)
n=nrow(df)
indexes = sample(n,n*(80/100))
trainset = df[indexes,]
testset = df[-indexes,]
dim(testset)
dim(trainset)
```

Specified the significant variables on the output variable at the level of 𝛼=0.05 and explored the related hypotheses test, estimating the parameters of the model.

```{r}
#Fits a GLM on the training set and displays the summary of the model.
dataglm = glm(y~.,trainset,family = "gaussian")
summary(dataglm)

```

### Predicting output of test dataset using trained model.  

**Functional form of the optimal predictive model.**

$$yhati=0.026+-0.011x_{1i}+0.001x_{2i}+-0.002x_{3i}$$

```{r}
#Predicts fatalities using the fitted GLM on the test set.
#Compares predictions ('pred') with actual values ('actual') and shows the first few predictions.

pred=predict(dataglm,testset)
actual=testset$y
values= data.frame(pred,actual)
head(values)
```

### RMSE

```{r}
#Calculates the Root Mean Square Error (RMSE) to evaluate the model's predictive accuracy.
RMSE=sqrt(sum(pred-actual)^2/length(actual))
RMSE

```
The Root Mean Square Error (RMSE) of the model demonstrates the model's high accuracy in predicting road fatalities

### Plotting Observed vs. Predicted Values

```{r}
# Predict values using the GLM model
predictions = predict(dataglm, type = "response")

# Plotting observed vs. predicted values
plot(actual, pred, xlab = "Observed Fatalities", ylab = "Predicted Fatalities", main = "Observed vs Predicted Fatalities")
abline(0, 1)

```

### Residuals Plot:

```{r}
# Calculating residuals
residuals = actual - pred

# Plotting residuals
plot(pred, residuals, xlab = "Predicted Fatalities", ylab = "Residuals", main = "Residuals vs Predicted Fatalities")

abline(h = 0)

```

### Diagnostic Plots:

```{r}
par(mfrow = c(2, 2))
plot(dataglm)

```

## Implications and Recommendations
+ The strong correlation between seatbelt usage and reduced fatalities underscores the importance of policies promoting seatbelt use.  
+ The negative correlation between income and fatalities suggests socio-economic factors may play a role in road safety.  
+ The GLM results highlight the significant impact of seatbelt usage, speed limits, and alcohol consumption on road fatalities. These insights can inform targeted road safety campaigns and policy decisions.  
+ Continued monitoring and analysis of these variables are recommended to further understand their long-term trends and impacts.

## Conclusion
The statistical analysis provided valuable insights into the factors affecting road safety. Seatbelt usage emerged as a key variable, along with income and age, in influencing road fatalities. The findings can guide future policy decisions and road safety measures to reduce fatalities effectively.